---
title: "latentsn: Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{latentsn: Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

```{r setup}
library(latentsn)
library(dplyr)
library(tidyr)
library(ggplot2)
```

This vignette shows the functionality available in the package `latentsn`. 

# Data input

Data should be a data frame of coordinates `x,y` object with a binary mark in variable `xii`. 

We demonstrate the package functionality on a synthetic data, which we now simulate. 

```{r}
set.seed(1)
# some parameters
p1 <- lsn_simpars_default(ndata = 300, 
                          # kernel shape
                          tau = .5,
                          # generator tree sizes, gamma-distr
                          psi_dbh = c(50, 200),
                          # Repulsion in the generator tree locations, Strauss
                          psi_lambda0 = 0.01,
                          psi_gamma = log(.1), # -inf hard-core, 0 none 
                          psi_range = 15, 
                          # logistic coefficients
                          theta = c(qlogis(.05), # ambient 5%
                                    log(2)) # effect of shotnoise.
                          )
# Simulate a random generator set, create the field, and mark.
sim1 <- lsn_simulate(p1)
```

For illustration, also compute the riskfield everywhere.

```{r}
U <- lsn_shotnoise(psi = sim1$psi, bbox = p1$bbox, resx = 2^8) |> 
  as_tibble()
risk <- mutate(U ,risk = plogis(p1$theta[1] + p1$theta[2] * p1$rho_scale * value))
```

Illustrate:

```{r}
grisk <- risk |> 
  ggplot() +
  # Latent generators
  geom_raster(aes(x=x, y=y, 
                  fill = risk
                  )
              ) +
  geom_point(data = sim1$psi, aes(x,y), shape = 3, col = "white") +
  # Observations 
  geom_point(data = sim1$phi, 
             aes(x = x, y = y, shape = infected, color = infected), 
             size = 2) + 
  #
  coord_fixed(exp=F, xlim = p1$bbox[1,], ylim = p1$bbox[2,]) +
  labs(fill = "P(infection)")  +
  scale_fill_distiller(palette = "Greys", limits = c(0,1), direction = 1)

grisk + ggtitle("True risk + data")
```

To get the shotnoise value at given locations, provide the coordinate(s) in the `grid` argument to `lsn_shotnoise`: 

```{r}
phi_u <- cbind(x = runif(50, 0, 100), y = runif(50, 0, 100)) 
U_u   <- lsn_shotnoise(psi = sim1$psi, grid = phi_u) 
glimpse(U_u)
```

# Estimating the model parameters

The model is estimated using a custom MCMC. To run many chains, use a wrapper.

```{r}
one_run <- function(...)
  lsn_mcmc_estimate(phi = sim1$phi, 
                    niter = 30000, 
                    keep_every = 50,
                    # priors
                    prior_lambda0 = c(100, 1000), # intensity of generators. make stiff
                    prior_psi = c(gamma = log(.1), range = 15), # repulsion amongst generators
                    prior_log_tau = c(log(.5), 0.2), # normal
                    prior_theta = list(m = c(plogis(.1), log(2)), S = diag(c(1, 1)) ), # mv-normal mean and covariance
                    print_every = Inf # disable output
  )

# then run in parallel with e.g. future-package. In this vignette we run 2 chains in a sequence.
mc1 <- one_run()
mc2 <- one_run()
mc <- list(mc1, mc2)
```

Some checks.

```{r}
# gather traces
tra <- lapply(1:2, \(i) mc[[i]]$hist_par |> mutate(iter = 1:n(), chain = i)) |> 
  bind_rows()
tral <- tra |> 
  pivot_longer(-c(iter, chain))
```


# Checks

Visual checks of convergence

```{r}
tral |> 
  ggplot() + 
  geom_line(aes(iter, value, col = factor(chain) )) +
  facet_wrap(~name, scale = "free", ncol = 3)
```

Make sure to run enough iterations. Note: The `U` parameter is the coefficient of the shot-noise field, and is estimated in **log-scale** with a log-normal prior.

Posterior densities

```{r}
tral |> 
  filter(iter > 5000 & iter %% 5 == 0) |> 
  ggplot() +
  geom_density(aes(value)) +
  facet_wrap(~name, scale = "free", ncol = 3)
```

# Posterior risks 

Samples from the posterior past-risk field need evaluation of the shotnoise field along the `keep_every` samples of the latent generators. The mcmc-object contains the matching iteration-index to match the kernel shape parameter, and the logistic parameters' samples. 

```{r}
idx <- mc1$hist_psi_it
i_keep <- which(idx > 15000 & idx %% 500 == 0)

Ul <- lapply(i_keep, \(i){
  # parameters
  pi <- mc1$hist_par[idx[i], ]
  tau <- exp(pi$log_tau)
  
  psii <- mc1$hist_psi[[i]] |> cbind( tau ) |> data.frame() |> setNames(nm = c("x","y","d", "tau"))
  Ui <- lsn_shotnoise(psi = psii, bbox = p1$bbox ) |> 
    as_tibble()
  # risk
  risk <- Ui |> 
    mutate(risk = plogis( pi$Intercept + p1$rho_scale * exp(pi$U) * value ),
           iter = i)
})

# Posterior mean risk
Upast <- bind_rows(Ul) |> 
  summarise(risk = mean(risk), .by = c(x, y))
```

Illustrate

```{r}
grisk %+% Upast + ggtitle("Posterior mean past risk estimate")
```

Similarly, future risk. Assume tree dbh .3m

```{r}
# fixed tree locations now
phi0 <- sim1$phi |>
  filter(xii == 1) |> 
  transmute(x, y, d = .3)

Uf <- lapply(i_keep, \(i){
  # parameters
  pi <- mc1$hist_par[idx[i], ]
  tau <- exp(pi$log_tau)
  psii <- phi0 |> mutate(tau = tau)
  Ui <- lsn_shotnoise(psi = psii, bbox = p1$bbox ) |> 
    as_tibble() 
  # risk
  risk <- Ui |> 
    mutate(risk = plogis( pi$Intercept + p1$rho_scale * exp(pi$U) * value ),
           iter = i)
})

# Posterior mean future infection risk
Ufuture <- bind_rows(Uf) |> 
  summarise(risk = mean(risk), .by = c(x, y))
```

Illustrate

```{r}
grisk %+% Ufuture + ggtitle("Posterior mean future risk estimate")
```

# Estimated kernel 

To check the kernel estimate, pull the samples and evaluate. Use again dbh=.3m.


```{r, fig.width=5, fig.height=4}
r <- seq(-1, 1, l = 101) * 10
taus <- exp(mc1$hist_par[i_keep, "log_tau"])
  
kernl <- lapply(taus, \(tau){
  tibble(r = r, kern = lsn_root_density_cutoff(abs(r), d = .3, tau = tau), i = tau)    
}) |> bind_rows()

kernl |> 
  ggplot() +
  geom_line(aes(r, kern, group = i), alpha = .1, linewidth = 2)
```

